Assumptions of Physics blueprint: information granularity as a common foundation for geometry, measures and probability

### 2021-03-30 Overviews/Informational granularity/main.tex
InfGran: review with Christine until family of measures

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}

\usepackage[margin=1in]{geometry}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
% Remove line spaces between items of enumerate and itemize
\usepackage{enumitem}
\setlist{noitemsep}
\usepackage{hyperref}
\hypersetup{
	colorlinks=true,
	citecolor=blue,
	urlcolor=blue,
	linkcolor=blue
}
\input{basicstyle}
\input{logic}
\input{theorems}

@@

\title{Assumptions of Physics blueprint: information granularity as a common foundation for geometry, measures and probability.}
\author{Gabriele Carcassi, Mark Greenfield, Christine Aidala }
\date{\today}

@@

@@

\begin{abstract}
    This paper presents ideas and progress towards an order theoretic framework that can be used as a common foundation for geometrical, measure theoretic and probabilistic concepts in the sciences. This would address foundational questions such as: under what assumptions standard tools like calculus are physically meaningful? How do geometrical structures arise in physics? How can the idea of physical dimension be properly captured mathematically? The working theory is that geometrical and measure theoretic structures capture one and only one thing: how coarse or fine grained are the statements within a scientific theory. They should  be recovered from a single binary operator, that allows to compare two statements and say whether one provides a finer level of description than the other. Mathematically, fineness imposes a preorder on the algebra of statements of a scientific theory. Measure theoretical and geometrical notions will be recovered by picking a unit and quantifying the degree of precision in terms of that unit: ``the mass of the object is between 1 and 3 Kg'' is twice less precise than ``the mass of the object is between 0 and 1 Kg''. The objective is to find necessary and/or sufficient conditions for such constructions which can be given a straight forward physical motivation.
%    As in comparative probability, we start with a $\sigma$-complete Boolean algebra that represents all possible descriptions in a given context and impose a preorder. In our case, the preorder does not represent likelyhood comparison, but rather whether one description is at a finer level of granularity than another. At this point, any statement (apart from an impossible one) can be chosen to act as a unit, and a measure (not necessarily bounded) can be constructed to quantify the degree to which the other statements give a more or less refined account. The starting lattice is not necessarily Archimedean, therefore the choice of unit will decide which elements are finite, infinite or infinitesimal. The family of measures so obtained are able to give a size to sets at different level of scale (e.g. points, lines, surfaces, ...) much like geometrical structures do. Our hope is that this structure recovers as special cases the geometrical, measure theoretical and probabilistic constructions we currently use in the sciences while giving a generalization that still works in the case where they fail (e.g. lack of $\sigma$-additivity in quantum probability) or expected to fail (e.g. breakdown at Planck's length).

@@

\emph{This paper presents the goal and the current status of our research. It is intended to gather early feedback, including pointers to other relevant work.}

The overall goal of Assumptions of physics is to to identify a handful of physical principles from which the basic laws can be rigorously derived. We have already established that topologies and $\sigma$-algebras can be recovered from requirements of experimental verifiability.\cite{AoPBook,AoPOverviewVerifiability} The next step is to understand what are the underpinning of geometrical and measure theoretic structures (including probability) which are layered on top of topologies and $\sigma$-algebras. As usual, a formal construction from scratch will force us to make explicit the physical requirements that are baked into the use of the common concepts and provide us with a framework that is, hopefully, still valid when those assumptions fail.

Note that often in physics geometric structures are recovered from other geometric structures through the idea of emergence. For example, substantial effort is put in constructing theories where space-time emerge from other more fundamental structures. This does not solve the problem: it simply pushes it backwards. We are interested in the more fundamental problem of how the first geometric structure, and in fact any geometric structure, is defined.

The idea is that while experimental verifiability keeps track of which statements in the theory are experimentally verifiable, informational granularity keeps track of which provides a finer level of description. \textbf{The working theory is that this, and only this, is what geometrical and measure theoretic structure capture: how coarse or fine grained are the statements within a scientific theory.} Finding the fundamental axioms associated to the notion of information granularity is therefore the central problem.

@@

Note that, conceptually, a statement is finer than another if it is true in fewer cases. Therefore the notion of fineness is linked to the notion of ``counting'' of the possible outcomes. It should be no surprise, then, that the same concept is compatible with the different disciplines above: geometrical size quantifies the ways we can arrange objects; probability theory quantifies the ratio of expected cases; information theory needs to quantify the different possible messages stored, computed or transmitted through an information system. Therefore the common structure we propose is not simply a mathematical pattern among disconnected structures: it is a single conceptual framework.

We found a few ideas from comparative probability useful, but our goal cannot be achieved with a single measure. A single measure can only compare objects with a finite measure. All objects with zero measure (or infinite measure) are indistinguishable. For example, we want to say:
\begin{itemize}
    \item $\stmt_1$ = \statement{The horizontal position of the object is exactly 0 meters}
    \item $\stmt_2$ = \statement{The horizontal position of the object is exactly 1 or 2 meters}
    \item $\stmt_3$ = \statement{The horizontal position of the object is between 0.5 and 1.5 meters}
    \item $\stmt_4$ = \statement{The horizontal position of the object is between 1.5 and 3.5 meters}
    \item $\stmt_1 \finer \stmt_2 \finer \stmt_3 \finer \stmt_4$
    \item $\stmt_1 \ncoarser \stmt_2 \ncoarser \stmt_3 \ncoarser \stmt_4$
\end{itemize}
Therefore we need to construct a family of measures, each identified by picking a particular statement that acts as a reference, as a unit. This provides a much richer structure, that is both interesting mathematically and more scientifically useful.

As for geometry, geometric structures provide measures for different dimensionality by starting from a lowest definition (i.e. line distance for Riemannian and surface are for symplectic) and then constructing higher dimensional measures through products. While want to recover those cases as well, this is not general enough. In physics there will be dimensions that are not directly comparable (e.g. pressure, volume, number of particles, chemical potential) but are still comparable under appropriate composition (i.e. pressure times volume with number of particles times chemical potential). We want to have a structure rich enough to capture these physical ideas (i.e. systems of units) which is not typically done.

@@

We start with a $\sigma$-complete Boolean algebra $\tdomain$, which we call \textbf{theoretical domain}, that can be generated from a countable subset, a countable basis. Conceptually, this represents all the statements that can be associated with an experimental verification procedure. The full detail of why this is the correct abstraction is discussed in detail in our previous work.\cite{AoPBook} It is a property of this type of spaces that each statement $\stmt \in \tdomain$ can be expressed as a disjunction of possible cases, of the  \textbf{possibilities} of the domain.\footnote{For example, ``the mass of the object is less than 1 Kg'' is the disjunction (i.e. the logical OR) of all the statements of the form ``the mass of the object is exact $x$ Kg'' where $0 \leq x < 1$.} Formally, each possibility is a minterm\footnote{A minterm is a disjunction in which each term appears once, either negated or not.} of the countable basis. The theoretical domain can then equivalently be represented by a $\sigma$-algebra.\footnote{This conceptually maps perfectly to what one has in probability: a sample space (i.e. the set of possibilities) and a set of events (i.e. the statements in the domain). The advantage of our ``pointless'' constructions is that we can treat domains, subdomains and composite domains in the same manner.}

@@

The whole program can be summarized in the following question: what intuitive conceptual properties does fineness have physically that, once properly formalized mathematically, allow us to recover measures, probability and geometrical structures?\footnote{In terms of the $\sigma$-algebra, how do we go from being able to compare the size of sets to being able to quantify it?}

@@

The first set of properties are the ones we must expect fineness to have in all cases. At this point, there are three, identified by DeFinetti\cite{definetti} for comparative probability, that we believe are necessary and that work as simple and intuitive starting points.

\textbf{Boundness}. As we mentioned in the introduction, a description is finer than another if it can be true in fewer cases. This intuitive notion is in general ill-defined except for two cases. A certainty $\certainty$ (e.g. \statement{the position is between minus infinity and plus infinity}) is true in all cases, therefore it must be the coarsest statement. An impossibility (e.g. \statement{the position is between 0 and 1 and also between 2 and 3}) is false in all cases, therefore it must be the finest statement. This means that fineness must be uniquely bound by the impossibility and the certainty. Any contingent statement (i.e. one that can in principle be either true or false) must be strictly between the two since there are some cases in which it will be true and others in which it will be false. This is the first required property.\footnote{In terms of the $\sigma$-algebra, the full set is strictly the biggest set and the empty set is strictly the smallest set.}

\textbf{Transitivity}. Additionally, if $\stmt_1$ gives finer description than $\stmt_2$ and $\stmt_2$ gives a finer description than $\stmt_3$ then it must be that $\stmt_1$ gives a finer description than $\stmt_3$. This means that fineness is transitive, and this is the second required property.

\textbf{Offset monotonicity}. Lastly, if we take two descriptions and we make them less precise ``in the same way'', we must have the same relationship in fineness. More precisely, suppose we have two descriptions $\stmt_1$ and $\stmt_2$ and we pick a third description $\stmt$ that is incompatible\footnote{Two statements are incompatible if they can't be both true at the same time: their conjunction $\stmt_1 \AND \stmt \equiv \impossibility$ is impossible. This is the analogue of disjoint sets in a $\sigma$-algebra.} with both. The respective disjunctions $\stmt_1 \OR \stmt$ and $\stmt_2 \OR \stmt$ will be respectively ``equally'' broader than the respective original statement. Therefore we will want them to be ``equally'' coarser as well. In other words, it must be that $\stmt_1 \finer \stmt_2$ if and only if $\stmt_1 \OR \stmt \finer \stmt_2 \OR \stmt$: fineness must be offset monotonic\footnote{Some authors call this property ``monotony'', which we don't find descriptive. We prefer ``offset monotonicity'' as it makes it clear that adding or removing a ``constant'' on both sides is a monotonic operation.} in the sense that it is invariant if we add or remove a statement from both sides, and this is the third and last required property.\footnote{In terms of the $\sigma$-algebra, a set is smaller than another if and only if it is also smaller when we increase both sets with another disjoint from the two.}

From these three properties, we can show the following:
\begin{description}
    \item[Preorder.] Fineness is a preorder over the theoretical domain $\tdomain$.
    \item[Monotonicity.] Fineness respects narrowness: $\stmt_1 \narrower \stmt_2$ implies $\stmt_1 \finer \stmt_2$.
    \item[Equigranularity is an equivalence.] Equigranularity, the symmetrization of fineness, is an equivalence class, and fineness is an order of those equivalence classes. That is, fineness orders the levels of description.
\end{description}

While we are not convinced these are sufficient properties to define fineness, we are also not found other properties that must be convincingly necessary. This is still an open area.

\subsubsection{Optional properties for fineness}

Optional properties for fineness include properties that are desirable, but are known to fail in some cases. These will be properties of a particular domain, rather than fineness itself.

\textbf{Equigranularity of possibilities}. Possibilities provide a complete characterization of a domain: if we know one possibility to be true (e.g. \statement{the mass of the photon is exactly 0 eV}), we know the truth value of all other statements (e.g. \statement{the mass of the photon is between x and y eV}). Therefore it is natural to assume that all possibilities are equigranular. In fact, this is a necessary property if we want to have a unique measure when we take a possibility as a unit.\footnote{The measure, in this case, is simply the counting measure and equigranularity of the possibilities plays a role similar to the principle of indifference in classical probability} However, this property can be easily violated by creating a subdomain.

@@

\textbf{Totality and physical dimensions}. If our objective is to construct a measure, totality is needed in the sense that every statement (i.e. every Borel set of possibilities) needs to be comparable with every other if we are to assign a unique number. However, statements whose size is defined in units of different physical dimensions (i.e. length, pressure, mass) should not be comparable. Lack of totality, in fact, should be a feature that allows us to formally construct a unit system.

For example, in phase space (i.e. symplectic manifold) we would expect points to be comparable and equigranular $\eqgran$: each state gives the same description. This means that we can compare sets of finite points. Finite areas are also comparable to each other (through integration of the symplectic form), and are comparable to points. However, vertical lines (i.e. ranges in momentum alone) are not comparable to horizontal lines (i.e. ranges in position alone). Symplectic geometry, in fact, gives a size to areas and not to lines.

Another counter example come from infinite ranges. Take the real line and consider the set of all positive numbers and the set of all negative numbers: should they be comparable? If we say they are equal, then that would tell us that there are more cases before 0 than after. This means that the space is not symmetrical under translation. If the real values represent distance between points, this is bad as it violates the idea that space is uniform. If the real values represents charge, however, this would actually make sense: the idea that there are exactly the same values for positive and negative charges does make sense, and the symmetry we are imposing is natural since zero is really a special number. Therefore different choices of comparability for the same Borel algebra is actually a feature we do not want to lose in general.

However, once we pick the unit, we should be able to find the ``right'' set of statements on which to create the measure. Roughly speaking, two description of the same units are ``finitely comparable'' in the sense that one gives a finer description of the other of a finite factor. Descriptions of different unites are either ``infinitely comparable'' (i.e. areas are always bigger than lengths) or not comparable (i.e. position and momentum). One approach would be to ask the ``largest set of statements mutually comparable in fineness that contain the unit''. In general, we have no expectation that this would be unique. So one question would be: when is this true?

@@

\textbf{Boolean lattice}. Granularity levels (i.e. equigranularity equivalence classes) form a bounded partially ordered set. One wonders whether they form a lattice. Physically this would make sense: we would be saying that given a set of statements $S \subseteq \tdomain$, we can always find a level of description that is the finest among all coarser descriptions than all the statements (i.e. $\bigveedot S$) and a level of description that is the coarsest among all finest descriptions than all the statements (i.e. $\bigwedgedot S$). However, it is not clear whether this is required, something only to be imposed in some cases or something that can be already proven.

Regardless, the map from a statement to its level of description (i.e. equivalence class in equigranularity), while it is an order isomorphism and preserves the top and bottom, is not a lattice isomorphism. That is, it does not preserve the meet and join. Let us note $\dot{\stmt} = {\stmt}_{/\eqgran}$ the level of granularity of $\stmt$. Now suppose that $\stmt_1 \eqgran \stmt_2 \nequiv \impossibility$ and $\stmt_1 \ncomp \stmt_2$. Then $\dot{\stmt}_1 = \dot{\stmt}_2 \neq \dot{\impossibility}$. Now suppose that $\stmt \to \dot{\stmt}$ is a lattice isomorphism. Then $\dot{\impossibility} = (\stmt_1 \AND \stmt_2)_{/\eqgran} = \dot{\stmt}_1 \wedgedot \dot{\stmt}_2 = \dot{\stmt}_1 \wedgedot \dot{\stmt}_1 = \dot{\stmt}_1$, which contradicts the premise.

\textbf{Addition}. We also need to understand how to define the addition and subtraction. Conceptually, the addition of two level of precision $\dot{\stmt}_1 + \dot{\stmt}_2$ means returning the level of precision that can be achieved by taking the disjunction $\stmt_1 \OR \stmt_2$ of two incompatible statements $\stmt_1 \in \dot{\stmt}_1$ and $\stmt_2 \in \dot{\stmt}_2$. This, of course, is not possible if the statements are ``bigger'' than ``half'' the space (e.g. $\stmt_1 \eqgran \stmt_2$, $\stmt_1 \comp \stmt_2$ and $\stmt_1 \OR \stmt_2 \equiv \certainty$). A more order theoretical definition would be to take all possible disjuctions of $\stmt[v]_1 \OR \stmt[v]_2$ such that $\stmt[v]_1 \in \dot{\stmt}_1$ and $\stmt[v]_2 \in \dot{\stmt}_2$ and return the coarsest level of description. That is, the sum is the least upper bound in fineness of all possible disjunctions. One should be able to prove that the incompatible disjunction corresponds to that level of description.

@@

\textbf{Missing axioms}. In many of our proofs, we ran into an issue of not being able to guarantee existence of certain statements. Several of the definitions we wanted included statements like ``say x holds if there exist statement y such that w." However, we have no axioms that guarantee any existence of any statements beyond $\certainty$ and $\impossibility$. In the case of an infinite collection of possibilities, we have no way to guarantee existence of statements of various levels of fineness when all statements in question are compatible with infinitely many possibilities.

To summarize, we may want to add something about existence of statements in equigranularity classes so as to have no ``gaps" (e.g. the order on $\tdomain/_{\eqgran}$ is complete). We may also want to say something about existence of enough statements representing each equivalence class so that sum and product can be well-defined. For example, if $\stmt_1\finer\stmt_2$, we want to say that there exists $\stmt_1'$ with $\stmt_1'\eqgran\stmt_1$ and $\stmt_1'\narrower\stmt_2$, as well as $\stmt_1''$ such that $\stmt_2\AND\stmt_1''$ is minimal in some sense ($\stmt_2\ncomp\stmt_1''$ if they do not fill the space). This is not at all guaranteed by the decomposition of statements into possibilities, since cardinality does not help us here.

\textbf{Bigger and smaller halfs}. One thing that appears to be missing is the following. Consider the following domain with four possibilities $\{x_1, x_2, x_3, x_4\}$. Each statement is the disjunction of a set of possibilities and let us use the lower indexes to indicate them. For example, $\stmt_{12} = x_1 \OR x_2$. Now, suppose $\stmt_{12} \broader \stmt_{34}$ and $\stmt_{14} \broader \stmt_{23}$. The two pairs are incompatible (disjoint) statements whose conjunction is the certainty. In a sense $\stmt_{14}$ is more than half the certainty while $\stmt_{34}$ is less than half the certainty. So, we would expect $\stmt_{14} \broader \stmt_{34}$. Yet, our axioms do not guarantee that. To prove that, we wrote a program that constructed all possible relationships given the axioms, and that relationship was not derived.

>>>>

\documentclass[10pt, onecolumn, nofootinbib]{revtex4-1}
\usepackage[utf8]{inputenc}

\usepackage[margin=1in]{geometry}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
% Remove line spaces between items of enumerate and itemize
\usepackage{enumitem}
\setlist{noitemsep}
\usepackage{hyperref}
\hypersetup{
	colorlinks=true,
	citecolor=blue,
	urlcolor=blue,
	linkcolor=blue
}
\input{basicstyle}
\input{logic}
\input{theorems}
\frenchspacing

@@

@@

	%TODO: add e-mail
	%TODO: not have page break on first page

\title{Assumptions of Physics blueprint: information granularity as a common foundation for geometry, measures and probability}
\author{Gabriele Carcassi, Mark Greenfield, Christine Aidala }
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}
\date{\today}

\begin{abstract}
    This paper presents ideas and progress towards an order theoretic framework that can be used as a common foundation for geometric, measure theoretic and probabilistic concepts in the sciences. This would address foundational questions such as: under what assumptions are standard tools like calculus physically meaningful? How do geometric structures arise in physics? How can the idea of physical dimension be properly captured mathematically? The working hypothesis is that geometric and measure theoretic structures capture one and only one thing: how coarse or fine grained the statements are within a scientific theory. Those structures should be recovered from a single binary operator, which allows us to compare two statements and say whether one provides a finer level of description than the other. Mathematically, fineness imposes a preorder on the algebra of statements of a scientific theory. Measure theoretic and geometric notions will be recovered by picking a unit and quantifying the degree of precision in terms of that unit: ``the mass of the object is between 1 and 3 Kg'' is half as precise as ``the mass of the object is between 0 and 1 Kg''. The objective is to find necessary and/or sufficient conditions for such constructions which can be given a straightforward physical motivation.
%    As in comparative probability, we start with a $\sigma$-complete Boolean algebra that represents all possible descriptions in a given context and impose a preorder. In our case, the preorder does not represent likelyhood comparison, but rather whether one description is at a finer level of granularity than another. At this point, any statement (apart from an impossible one) can be chosen to act as a unit, and a measure (not necessarily bounded) can be constructed to quantify the degree to which the other statements give a more or less refined account. The starting lattice is not necessarily Archimedean, therefore the choice of unit will decide which elements are finite, infinite or infinitesimal. The family of measures so obtained are able to give a size to sets at different level of scale (e.g. points, lines, surfaces, ...) much like geometrical structures do. Our hope is that this structure recovers as special cases the geometrical, measure theoretical and probabilistic constructions we currently use in the sciences while giving a generalization that still works in the case where they fail (e.g. lack of $\sigma$-additivity in quantum probability) or expected to fail (e.g. breakdown at Planck's length).

@@

@@

\emph{This paper presents the goal and the current status of our research. It is intended to gather early feedback, including pointers to other relevant work. Feedback is welcome and encouraged.}

The overall goal of Assumptions of Physics is to identify a handful of physical principles from which the basic laws can be rigorously derived. We have already established that topologies and $\sigma$-algebras can be recovered from requirements of experimental verifiability.\cite{AoPBook,AoPOverviewVerifiability} See also (TODO: Kevin Kelly). The next step is to understand the underpinnings of geometric and measure theoretic structures (including probability) which are layered on top of topologies and $\sigma$-algebras. As usual, a formal construction from scratch will force us to make explicit the physical requirements that are baked into the use of the common concepts and provide us with a framework that is, hopefully, still valid when those assumptions fail.
%TODO: add topology proceeding, update book reference

Note that often in physics, geometric structures are recovered from other geometric structures through the idea of emergence. For example, substantial effort is put into constructing theories where space-time emerges from other more fundamental structures. This does not solve the problem: it simply pushes it backwards. We are interested in the more fundamental problem of how the first geometric structure, and in fact any geometric structure, is defined.

The idea is that while experimental verifiability keeps track of which statements in the theory are experimentally verifiable, informational granularity keeps track of which provide a finer level of description. \textbf{The working hypothesis is that this, and only this, is what geometric and measure theoretic structures capture: how coarse or fine grained the statements are within a scientific theory.} Finding the fundamental axioms associated to the notion of information granularity is therefore the central problem.

@@

Note that, conceptually, a statement is finer than another if it is true in fewer cases. Therefore the notion of fineness is linked to the notion of ``counting'' of the possible outcomes. It should be no surprise, then, that the same concept is compatible with the different disciplines above: geometric size quantifies the ways we can arrange objects; probability theory quantifies the ratio of expected cases; information theory quantifies the different possible messages stored, computed or transmitted through an information system. Therefore the common structure we propose is not simply a mathematical pattern among disconnected structures: it is a single conceptual framework.

We found a few ideas from comparative probability useful, but our goal cannot be achieved with a single measure. A single measure can only compare objects of finite measure. All objects with zero measure (or infinite measure) are indistinguishable. For example, we want to say:
\begin{itemize}
    \item $\stmt_1$ = \statement{The horizontal position of the object is exactly 0 meters}
    \item $\stmt_2$ = \statement{The horizontal position of the object is exactly 1 or 2 meters}
    \item $\stmt_3$ = \statement{The horizontal position of the object is between 0.5 and 1.5 meters}
    \item $\stmt_4$ = \statement{The horizontal position of the object is between 1.5 and 3.5 meters}
    \item $\stmt_1 \finer \stmt_2 \finer \stmt_3 \finer \stmt_4$
    \item $\stmt_1 \ncoarser \stmt_2 \ncoarser \stmt_3 \ncoarser \stmt_4$
\end{itemize}
Therefore we need to construct a family of measures, each identified by picking a particular statement that acts as a reference, as a unit. This provides a much richer structure, which is both interesting mathematically and more scientifically useful.

As for geometry, geometric structures provide measures for different dimensionality by starting from a lowest dimension (i.e. line distance for Riemannian geometry and surface area for symplectic geometry) and then constructing higher dimensional measures through products. While we want to recover those cases as well, this is not general enough. In physics there will be dimensions that are not directly comparable (e.g. pressure, volume, number of particles, chemical potential) but are still comparable under appropriate composition (i.e. pressure times volume with number of particles times chemical potential). We want to have a structure rich enough to capture these physical ideas (i.e. systems of units) which is not typically done.

@@

We start with a $\sigma$-complete Boolean algebra $\tdomain$, which we call \textbf{theoretical domain}, that can be generated from a countable subset, a countable basis. Conceptually, this represents all the statements that can be associated with an experimental verification procedure. The full detail of why this is the correct abstraction is discussed in detail in our previous work.\cite{AoPBook} It is a property of this type of spaces that each statement $\stmt \in \tdomain$ can be expressed as a disjunction of possible cases, of the  \textbf{possibilities} of the domain.\footnote{For example, ``the mass of the object is less than 1 Kg'' is the disjunction (i.e. the logical OR) of all the statements of the form ``the mass of the object is exactly $x$ Kg'' where $0 \leq x < 1$.} Formally, each possibility is a minterm\footnote{A minterm is a disjunction in which each term appears once, either negated or not.} of the countable basis. The theoretical domain can then equivalently be represented by a $\sigma$-algebra.\footnote{This conceptually maps perfectly to what one has in probability: a sample space (i.e. the set of possibilities) and a set of events (i.e. the statements in the domain). The advantage of our ``pointless'' constructions is that we can treat domains, subdomains and composite domains in the same manner.}

@@

The whole program can be summarized in the following question: what intuitive conceptual properties does fineness have physically that, once properly formalized mathematically, allow us to recover measures, probability and geometric structures?\footnote{In terms of the $\sigma$-algebra, how do we go from being able to compare the size of sets to being able to quantify it?}

@@

The first properties are the ones we must expect fineness to have in all cases. At this point, there are three, identified by DeFinetti\cite{definetti} for comparative probability, that we believe are necessary and that work as simple and intuitive starting points.

\textbf{Boundedness}. As we mentioned in the introduction, a description is finer than another if it can be true in fewer cases. This intuitive notion is in general ill-defined except for two cases. A certainty $\certainty$ (e.g. \statement{the position is between minus infinity and plus infinity}) is true in all cases, therefore it must be the coarsest statement. An impossibility (e.g. \statement{the position is between 0 and 1 and also between 2 and 3}) is false in all cases, therefore it must be the finest statement. This means that fineness must be uniquely bound by the impossibility and the certainty. Any contingent statement (i.e. one that can in principle be either true or false) must be strictly between the two since there are some cases in which it will be true and others in which it will be false. This is the first required property.\footnote{In terms of the $\sigma$-algebra, the full set is strictly the biggest set and the empty set is strictly the smallest set.}

\textbf{Transitivity}. Additionally, if $\stmt_1$ gives a finer description than $\stmt_2$ and $\stmt_2$ gives a finer description than $\stmt_3$ then it must be that $\stmt_1$ gives a finer description than $\stmt_3$. This means that fineness is transitive, and this is the second required property.

\textbf{Offset monotonicity}. Lastly, if we take two descriptions and we make them less precise ``in the same way'', we must have the same relationship in fineness. More precisely, suppose we have two descriptions $\stmt_1$ and $\stmt_2$ and we pick a third description $\stmt$ that is incompatible\footnote{Two statements are incompatible if they can't both be true at the same time: their conjunction $\stmt_1 \AND \stmt \equiv \impossibility$ is impossible. This is the analogue of disjoint sets in a $\sigma$-algebra.} with both (noted $\stmt \ncomp \stmt_1$ and $\stmt \ncomp \stmt_2$). The respective disjunctions $\stmt_1 \OR \stmt$ and $\stmt_2 \OR \stmt$ will be ``equally'' broader than the original statements. Therefore we will want them to be ``equally'' coarser as well. In other words, it must be that $\stmt_1 \finer \stmt_2$ if and only if $\stmt_1 \OR \stmt \finer \stmt_2 \OR \stmt$: fineness must be offset monotonic\footnote{Some authors call this property ``monotony'', which we don't find descriptive. We prefer ``offset monotonicity'' as it makes it clear that adding or removing a ``constant'' on both sides is a monotonic operation.} in the sense that it is invariant if we add or remove a statement from both sides, and this is the third and last required property.\footnote{In terms of the $\sigma$-algebra, a set is smaller than another if and only if it is also smaller when we increase both sets with another disjoint from the two.}

From these three properties, we can show the following:
\begin{description}
    \item[Preorder] Fineness is a preorder over the theoretical domain $\tdomain$.
    \item[Monotonicity] Fineness respects narrowness: $\stmt_1 \narrower \stmt_2$ implies $\stmt_1 \finer \stmt_2$.
    \item[Equigranularity is an equivalence] Equigranularity $\eqgran$, the symmetrization of fineness, is an equivalence, and fineness is a partial order of its equivalence classes. That is, fineness orders the levels of description.
\end{description}

While we are not convinced these are sufficient properties to define fineness, we have not found other properties that we believe to be necessary. This is still an open area.

\subsubsection{Domain-specific properties for fineness}

Domain-specific properties for fineness include properties that are desirable, but are known to fail in some cases. These will be properties of a particular domain, rather than fineness itself.

\textbf{Equigranularity of possibilities}. Possibilities provide a complete characterization of a domain: if we know one possibility to be true (e.g. \statement{the mass of the photon is exactly 0 eV}), we know the truth value of all other statements (e.g. \statement{the mass of the photon is between x and y eV}). Therefore it is natural to assume that all possibilities are equigranular. In fact, this is a necessary property if we want to have a unique measure when we take a possibility as a unit.\footnote{The measure, in this case, is simply the counting measure and equigranularity of the possibilities plays a role similar to the principle of indifference in classical probability.} However, this property can be easily violated by creating a subdomain.

@@

\textbf{Totality and physical dimensions}. If our objective is to construct a measure, totality is needed in the sense that every statement (i.e. every Borel set of possibilities) needs to be comparable with every other if we are to assign a unique number. However, statements whose size is defined in units of different physical dimensions (e.g. length, pressure, mass) should not be comparable. Lack of totality, in fact, should be a feature that allows us to formally construct a unit system.

For example, in phase space (i.e. in a symplectic manifold) we would expect points to be comparable and equigranular: each state gives the same description. This means that we can compare sets of finite points. Finite areas are also comparable to each other (through integration of the symplectic form), and are comparable to points: they will be infinitely coarser. However, vertical lines (i.e. ranges in momentum alone) are not comparable to horizontal lines (i.e. ranges in position alone). Symplectic geometry, in fact, gives a size to areas and not to lines.

Another counterexample comes from infinite ranges. Take the real line and consider the set of all positive numbers and the set of all negative numbers: should they be comparable? If we say they are equal, then that would tell us that there are the same number of cases before and after 0. This means that the space is not symmetric under translation. If the real values represent distance between points, this is bad as it violates the idea that space is uniform. If the real values represent charge, however, this would actually make sense: the idea that there are exactly the same values for positive and negative charges does make sense, and the symmetry we are imposing is natural since zero is really a special number. Therefore different choices of comparability for the same Borel algebra is actually a feature we do not want to lose in general.

However, once we pick the unit, we should be able to find the ``right'' set of statements on which to create the measure. Roughly speaking, two descriptions of the same units are ``finitely comparable'' in the sense that one gives a finer description than the other by a finite factor. Descriptions of different units are either ``infinitely comparable'' (e.g. areas are always bigger than lengths) or not comparable (e.g. position and momentum). One approach would be to find the ``largest set of statements mutually comparable in fineness that contain the unit''. In general, we have no expectation that this would be unique. So one question would be: when is this set unique?

@@

\textbf{Boolean lattice}. Granularity levels (i.e. equigranularity equivalence classes) form a bounded partially ordered set. One wonders whether they form a lattice. Physically this would make sense: we would be saying that given a set of statements $S \subseteq \tdomain$, we can always find a level of description that is the finest among all coarser descriptions than all the statements (i.e. $\bigveedot S$) and a level of description that is the coarsest among all finer descriptions than all the statements (i.e. $\bigwedgedot S$). However, it is not clear whether this is required, something only to be imposed in some cases or something that can be already proven.

Regardless, the map from a statement to its level of description (i.e. equivalence class in equigranularity), while it is an order isomorphism and preserves the top and bottom, is not a lattice isomorphism. That is, it does not preserve the meet and join. Let us note $\dot{\stmt} = {\stmt}_{/\eqgran}$ the level of granularity of $\stmt$. Now suppose that $\stmt_1 \eqgran \stmt_2 \nequiv \impossibility$ and $\stmt_1 \ncomp \stmt_2$ (recall $\ncomp$ means incompatible). Then $\dot{\stmt}_1 = \dot{\stmt}_2 \neq \dot{\impossibility}$. Now suppose that $\stmt \to \dot{\stmt}$ is a lattice isomorphism. Then $\dot{\impossibility} = (\stmt_1 \AND \stmt_2)_{/\eqgran} = \dot{\stmt}_1 \wedgedot \dot{\stmt}_2 = \dot{\stmt}_1 \wedgedot \dot{\stmt}_1 = \dot{\stmt}_1$, which contradicts the premise.

\textbf{Addition}. We also need to understand how to define the addition and subtraction. Conceptually, the addition of two granularity levels $\dot{\stmt}_1 + \dot{\stmt}_2$ means returning the granularity level that can be achieved by taking the disjunction $\stmt_1 \OR \stmt_2$ of two incompatible statements $\stmt_1 \in \dot{\stmt}_1$ and $\stmt_2 \in \dot{\stmt}_2$. This, of course, is not possible if the statements are ``bigger'' than ``half'' the space (e.g. $\stmt_1 \eqgran \stmt_2$, $\stmt_1 \comp \stmt_2$ and $\stmt_1 \OR \stmt_2 \equiv \certainty$). A more order theoretic definition would be to take all possible disjuctions of $\stmt[v]_1 \OR \stmt[v]_2$ such that $\stmt[v]_1 \in \dot{\stmt}_1$ and $\stmt[v]_2 \in \dot{\stmt}_2$ and return the coarsest level of description. That is, the sum is the least upper bound in fineness of all possible disjunctions. One should be able to prove that the incompatible disjunction corresponds to that level of description.

@@

\textbf{Missing axioms}. In many of our proofs, we ran into an issue of not being able to guarantee existence of certain statements. Several of the definitions we want have the form ``x holds if there exists statement y such that w." However, we have no axioms that guarantee the existence of any statements beyond $\certainty$ and $\impossibility$. In the case of an infinite collection of possibilities, we have no way to guarantee existence of statements of various levels of fineness when all statements in question are compatible with infinitely many possibilities.

To summarize, we may want to add something about existence of statements in equigranularity classes so as to have no ``gaps" (e.g. the order on $\tdomain/_{\eqgran}$ is complete). We may also want to say something about existence of enough statements representing each equivalence class so that sum and product can be well-defined. For example, if $\stmt_1\finer\stmt_2$, we may want to say that there exists $\stmt_1'$ with $\stmt_1'\eqgran\stmt_1$ and $\stmt_1'\narrower\stmt_2$, as well as $\stmt_1''$ such that $\stmt_2\AND\stmt_1''$ is minimal in some sense ($\stmt_2\ncomp\stmt_1''$ if they do not fill the space). This is not at all guaranteed by the decomposition of statements into possibilities, since cardinality does not help us here.

\textbf{Bigger and smaller ``halves''}. One thing that appears to be missing is the following. Consider the following domain with four possibilities $\{x_1, x_2, x_3, x_4\}$. Each statement is the disjunction of a set of possibilities and let us use the lower indices to indicate them. For example, $\stmt_{12} = x_1 \OR x_2$. Now, suppose $\stmt_{12} \coarser \stmt_{34}$ and $\stmt_{14} \coarser \stmt_{23}$. The two pairs are incompatible (disjoint) statements whose conjunction is the certainty. In a sense $\stmt_{14}$ is more than half the certainty while $\stmt_{34}$ is less than half the certainty. So, we would expect $\stmt_{14} \coarser \stmt_{34}$. Yet, our axioms do not guarantee that. To prove that, we wrote a program that constructed all possible relationships given the axioms, and that relationship was not derived.
